Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-07-24 Manuscript/MPESubmission/Revision/response.tex

\end{enumerate}

\section{Reviewer 2:}

\begin{enumerate}
\item{\textcolor{blue}{\textbf{Compare the trees to the "true" tree:} I don't see how the "best" tree approach is helpful in this case. It adds an additional layer of complexity that can't be accounted for via simulation. The point of simulations is to compare empirical realities to the generating process, to see how much results diverge. I think the most helpful and straightforward thing to do is eliminate the "best" tree part of the MS, and compare the inferred trees directly to the real topology to quantify variance.}}

A similar point/suggestion was also raised by reviewer 1 (see Reviewer 1 General comment number 6 and specific comment number 25).
We do understand the concerns of this reviewer as well but, again, we do not agree that comparing the "missing-data" trees to the "best" trees adds a layer of complexity to our simulations.
In fact, we argue that doing so actually adds a layer of complexity to the analysis.
The problematic we are tackling in this manuscript is to understand "what is the effect of our three parameters on recovering topology ?"
To answer this question, however, we have to add a layer between the data (i.e. the "complete" or "missing-data" matrices) and the actual topology which is the tree inference.
We test the effect of this layer by using different tree inference methods.

Because we use a simulation approach to test this problematic, we have to start from a random starting point (the "true" tree) and then do the following steps: (1) generate and remove the data (2) infer the topologies and (3) compare the topologies.
Note that in this schematic representation, the "true" tree is the step (0) of our analysis and involves a series of algorithms and softwares (i.e. seqGen\{phyclust\}, rTraitDisc\{ape\}, RAxML, MrBayes).
To compare the "true" tree's topology to the "missing-data" topologies will be equivalent to explore the effect of our three parameters on recovering topology through 3 layers (1- generate the data; 2- infer the topologies; and 3- compare the topologies).
By comparing the "best" tree's topology to the "missing-data" ones, we explore the effect of our three parameters through only 2 layers (2- infer the topologies; and 3- compare the topologies).
Note that we explore the effect of these 2 layers by using two different tree inference methods (effectively 4 since we also use the Maximum Likelihood bootstraps trees and the Bayesian posterior trees) and 2 different Normalised topological metrics (Robinson-Foulds and Triplets).
Using as main results the comparisons between the "true" and the "missing-data" trees would also need the exploration of the effect of the cascade of algorithms and softwares (maybe by modifying the algorithms used to generate the "true" tree and the matrices?).

However, we have added all the comparisons between the "true" tree and the "missing-data" trees in the supplementary.
It is interesting to notice that the algorithms and softwares as well as probably the size of our matrices are actually not ideal an the recovery of the "true" tree is rather low and scales down the our results (see figure @).
However, even though these comparisons are worst, it is encouraging to see that the patterns are conserved (i.e. Bayesian consensus trees $>$ Maximum Likelihood trees $>$ both bootstraps trees and Bayesian posterior trees).

### 2015-07-27 Manuscript/MPESubmission/Revision/response.tex

However, we have added all the comparisons between the "true" tree and the "missing-data" trees in the supplementary.
It is interesting to notice that the algorithms and softwares as well as probably the size of our matrices are actually not ideal an the recovery of the "true" tree is rather low and scales down the our results (see figure @).
However, even though these comparisons are worst, it is encouraging to see that the patterns are conserved (i.e. Bayesian consensus trees $>$ Maximum Likelihood trees $>$ both bootstraps trees and Bayesian posterior trees).

\item{\textcolor{blue}{\textbf{Better explanation of the missing data:} I found the explanation of the missing data strategy very difficult to follow. First, they oddly use the verb "to infer" for matrices; are they inferring matrices? Second, it's not clear anywhere that the authors actually removed the molecular data from the fossil taxa, though I assume they did. I hope the authors can re-write in simpler language ("explain like I'm 5") for easier comprehension.}}

The lines 180 and 181 states "All the molecular information for fossil taxa was replaced by missing data ("?")" in the former manuscript. We changed the verb "infer" (see specific comment number 16 from reviewer 1).

We assumed that this reviewer is mentioning the section "2.2 Removing data" as the "explanation of the missing data strategy". We modified the whole section emphasising on clarity and simplicity.

"To explore the effect of missing data on topological recovery, we removed various amounts of the “complete” matrix to obtain matrices with missing morphological data.Hereafter, we call these matrices with missing morphological data the “missing-data” matrices. We removed morphological following three main data incompleteness parameters:
1. The proportion of missing living taxa ($M_L$). This first missing-data parameter corresponds to the proportion of living taxa with no morphological data. It represents the number of living taxa that are present in the matrix but have only molecular data available. This reflects the fact that, because of the increasing facility of collecting molecular data, morphological data for living species is rarely collected (Guillerme and Cooper, 2015). Therefore, many living species will have only molecular data available. In practice, we removed all the morphological data from randomly chosen living with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
2. The proportion of missing data in the fossil record ($M_F$). This missing data parameter represents the completeness of the fossil record. In fact, due to preservation biases, missing data for fossil taxa are common (Sansom and Wills, 2013). In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
3. The number of morphological characters for both living and fossil taxa ($N_C$). This parameter is not a missing data parameter per se but rather an indication of the size of the matrix. Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. Therefore, this parameter correspond to the overall number of characters available for both living and fossil taxa. In practice, we randomly removed whole characters from the morphological matrix shrinking it to: 100, 90,75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of “missing” morphological characters).

In practice, each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living taxa’s data; $M_F$ removes cells from the fossil taxa’s data; and $N_C$ removes columns across both living and fossil taxa’s data. Note that $M_L$  and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

We created matrices using all parameter combinations resulting in 125 ($5^3$) “missing-data” matrices. Note that one of these combinations ($M_L$=0\%; $M_F$=0\% and $N_C$=100) has no missing data so is equivalent to the “complete” matrix, thus we have one effectively complete matrix in our 125 “missing-data” matrices. In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters. To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon. Note that the living taxa always had at least 90\% of data (the 1000 molecular characters)." (line @@@)

@@

\item{\textcolor{blue}{\textbf{$M_C$ is not missing data per se:} If Mc is just removing columns, then they are shrinking the total matrix. This is not missing data per se. They are instead comparing matrices of different sizes, and the quantity is "data present." Any matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. A matrix with 10,000 characters is tiny if 10,000,000 could be scored. This is semantic.}}

% TG: Ok i've changed it every where but personally I don't like it. It makes it even more messy and hard to read. However, because I didn't answered the first and last comment of this reviewer, I felt that at least 2/4 is a minimum.
We entirely agree with this comment and hope it will help to the global comprehension of the manuscript. We replaced the $M_C$ (missing characters) parameter by $N_C$ (number of characters with five states: 100, 90, 75, 50 and 25) in the revised manuscript (see reviewer 2 comment 2 above), the appendices and the figures. We also slightly modified some explanations related to the change of missing characters to number of characters:

### 2015-07-27 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Response edit

\subsection{Removing data}
\label{Removing_data}
We modified the ``complete'' matrix to obtain matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
\begin{enumerate}
\item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
\item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
\item{$N_{C}$, the number of morphological characters for both living and fossil taxa: 100, 90, 75, 50 or 25 characters. }
This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters \citep[e.g][]{O'Leary08022013}.
\end{enumerate}

In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $N_C$ removes columns across both living and fossil taxa's data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.
We first applied the parameters $M_L$ and $M_F$ to the matrix and then applied the $N_C$ parameter.
Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix.
When applying the $N_C$ parameter with the second parameter state (90 characters), however, the resulting matrix potentially had more than 10\% data missing.
Not that the $N_C$ parameter states can be seen in the same way as the $M_L$ and $M_F$ ones: the first state (100 characters) corresponds to 0\% less characters than the ``complete'' matrix, the second (90 characters), corresponds to 10\% less than the ``complete'' matrix, etc.

We created matrices using all parameter combinations resulting in 125 ($5^3$) matrices.
Note that one of these combinations has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.

>>>>

\subsection{Removing data}
% \label{Removing_data}
% We modified the ``complete'' matrix to obtain matrices with missing data by randomly replacing data with ``?" in the morphological part of the matrices according to the following parameters:
% \begin{enumerate}
% \item{$M_{L}$, the proportion of living taxa with no morphological data: 0\%, 10\%, 25\%, 50\% or 75\%.}
% This parameter illustrates the number of living taxa that are present in the molecular part of the matrix but not in the morphological part. This reflects the fact that the increased availability of molecular data means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis.
% \item{$M_{F}$, the proportion of missing data in the fossil taxa: 0\%, 10\%, 25\%, 50\% or 75\%.}
% This parameter illustrates the quality of the fossil record, a common problem due to preservation biases \citep{sansomfossilization2013}. 
% \item{$N_{C}$, the number of morphological characters for both living and fossil taxa: 100, 90, 75, 50 or 25 characters. }
% This parameter illustrates the number of available morphological characters for both living and fossil taxa. This reflects the amount of effort put into identifying the characters \citep[e.g][]{O'Leary08022013}.
% \end{enumerate}

% In practice, each parameter represents a different way of removing data from the matrix: $M_L$ removes rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $N_C$ removes columns across both living and fossil taxa's data.
% Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.
% We first applied the parameters $M_L$ and $M_F$ to the matrix and then applied the $N_C$ parameter.
% Therefore, when 10\% data was missing for both $M_L$ and $M_F$, 10\% data was missing in the morphological part of the matrix.
% When applying the $N_C$ parameter with the second parameter state (90 characters), however, the resulting matrix potentially had more than 10\% data missing.

% We created matrices using all parameter combinations resulting in 125 ($5^3$) matrices.
% Note that one of these combinations has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
% To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.

% TG: this part is entirely re-written.
To explore the effect of missing data on topological recovery, we removed various amounts of the ``complete'' matrix to obtain matrices with missing morphological data.
Hereafter, we call these matrices with missing morphological data the ``missing-data'' matrices.
We removed morphological following three main data incompleteness parameters:
\begin{enumerate}
\item{The proportion of missing living taxa ($M_{L}$).}
This first missing-data parameter corresponds to the proportion of living taxa with no morphological data.
It represents the number of living taxa that are present in the matrix but have only molecular data available.
This reflects the fact that, because of the increasing facility of collecting molecular data, morphological data for living species is rarely collected \citep{GuillermeCooperMissing}. % TG: see what I did here? ;)
Therefore, many living species will have only molecular data available.
In practice, we removed all the morphological data from randomly chosen living with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of living taxa with no morphological data.
\item{The proportion of missing data in the fossil record ($M_{F}$).}
This missing data parameter represents the completeness of the fossil record.
In fact, due to preservation biases, missing data for fossil taxa are common \citep{sansomfossilization2013}. 
In practice, we randomly removed a proportion of data among the fossil taxa with five different proportions: 0\%, 10\%, 25\%, 50\% or 75\% of overall missing data for the fossil taxa.
\item{The number of morphological characters for both living and fossil taxa ($N_{C}$).}
This parameter is not a missing data parameter \textit{per se} but rather an indication of the size of the matrix.
Any morphological matrix of any size has indeterminate missing data, given that the total number of characters is undefined, but presumably large. % TG: this is word for word what the reviewer wrote. I didn't like doing that first but after seeing the Ineterface Focus review and how they copy pasted the discussion suggestion, I think it's really nice for the reviewer to see a bit of it's work directly in the paper (at least it was for me!).
Therefore, this parameter correspond to the overall number of characters available for both living and fossil taxa.
In practice, we randomly removed whole characters from the morphological matrix shrinking it to: 100, 90, 75, 50 or 25 characters.
Note that these levels are equivalent to the two other parameters (i.e. 0\%, 10\%, 25\%, 50\% or 75\% of ``missing'' morphological characters).
\end{enumerate}

% TG: this part is nearly exactly the same as before.
In practice, each parameter represents a different way of removing data from the morphological part of the matrix: $M_L$ removes entire rows from the living taxa's data; $M_F$ removes cells from the fossil taxa's data; and $N_C$ removes columns across both living and fossil taxa's data.
Note that $M_L$ and $M_F$ differ not only because of the region of the matrix affected: for $M_L$ all the morphological data of a percentage of living taxa are removed, whereas for $M_F$ a percentage of the data are removed at random from across the whole of the morphological matrix for fossil taxa.

% TG: and this one is slightly modified only.
We created matrices using all parameter combinations resulting in 125 ($5^3$) ``missing-data''matrices.
Note that one of these combinations ($M_{L}$=0\%; $M_{F}$=0\% and $N_{C}$=100) has no missing data so is equivalent to the ``complete'' matrix, thus we have one effectively complete matrix in our 125 ``missing-data'' matrices.
In practice, we first removed the data following the two missing data parameters $M_L$ and $M_F$ and then removed data following the $N_C$ parameters.
To avoid avoid matrices containing taxa without any data (morphological or molecular), we repeated the random deletion until the matrices contained at least 5\% of data for any taxon.
Note that the living taxa always had at least 90\% of data (the 1000 molecular characters).
